\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-07-08 phd-cs-us.tex
updates

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size of the committee depends on the department size, e.g., around 15 at GMU. Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can provide inputs and opinions on your profile.

@@

It's important to aim for the top venues in your field, which you can
find on places such as CSRankings~\cite{csrankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences}. Local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty.

@@

\begin{commentbox}[Vu]
Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1 and. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.47\textwidth}
    \vspace{-0.35in}
      \begin{center}
        \includegraphics[width=0.47\textwidth]{c6.png}
      \end{center}
    \vspace{-0.4in}
    \end{wrapfigure}
Many students have letters written by the themselves and signed by their professors. These have little
values. We can easily recognize them and will consider them weakness.
Similarly, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.

@@

While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

\begin{wrapfigure}{r}{0.52\textwidth}
\vspace{-0.35in}
  \begin{center}
    \includegraphics[width=0.52\textwidth]{c2.png}
  \end{center}
\vspace{-0.4in}  
\end{wrapfigure}
Finally, this is something easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? provide names of professors who you're interested in (in many cases your application will be forwarded to them for evaluation).
This shows that you're serious and have done homework on places you're applying to.
Admission committee will look for this part at the end of the statement, so do not skip it.

@@

So what can you do about this? several things such as asking your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school) and explaining about your school in your statement (and asking your LOR writer to do that too). Of course, if you're interested in working with Vietnamese professors, consider a PhD program in the US that  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{have them}.

\begin{commentbox}[Vu]
Sometime PhD admission committee will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. In some cases we ask people (faculty, students) if we think they know about the place.  For example, when I was a postdoc at UMD, people in admission committee ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}

@@

@@

\begin{wrapfigure}{r}{0.48\textwidth}
    \vspace{-0.3in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
    \vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \vspace{-0.2in}
    \end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\begin{wrapfigure}{l}{0.5\textwidth}
    \vspace{-0.4in}
      \begin{center}
        \includegraphics[width=0.5\textwidth]{c5.png}
      \end{center}
    \vspace{-0.9in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

>>>>

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note that at GMU (and probably many other schools) we have a separate committee for the MS program.

Each application is then assigned to about \emph{three} faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can provide inputs and opinions on your profile.

@@

It's important to aim for the top venues in your field. You can
find top CS conferences at places such as CSRankings~\cite{csrankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences}. Local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty.

@@

\begin{commentbox}[Vu]
Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.47\linewidth}
      \begin{center}
        \includegraphics[width=1\linewidth]{c6.png}
      \end{center}
\end{wrapfigure}
Many students have letters written by themselves and signed by their professors. These have little
values. We can recognize them and will consider them weakness.
Similarly, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.

@@

\begin{wrapfigure}{r}{0.52\linewidth}
    \begin{center}
      \includegraphics[width=1\linewidth]{c2.png}
    \end{center}
  \end{wrapfigure}
While you might not have control over LORs \hyperref[sec:your-school]{where your go to school}, you do over your
statement! So write it well because we take it seriously.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

Finally, this is something easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? provide names of professors who you're interested in (in many cases your application will be forwarded to them for evaluation).
This shows that you're serious and have done homework on places you're applying to.
Admission committee will look for this part at the end of the statement, so do not skip it.

@@

So what can you do about this? several things including asking your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school) and explaining about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese professors, consider a PhD program in the US that  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{have them}.

\begin{commentbox}[Vu]
Sometime PhD admission committee in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. In some cases we ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, people in admission committee ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}

@@

\begin{commentbox}[Vu]
Personally I do not pay attention to GRE scores as I do not believe they are good indicator for research. However, if you have good GRE scores then you should list them in your application.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.48\textwidth}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
    \end{wrapfigure}
The short answer is that tenured-track faculty (e.g., assistant professors) are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding as tenured faculty.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
      \begin{center}
        \includegraphics[width=0.28\textwidth]{c4a.png}
      \end{center}
    \end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\begin{wrapfigure}{l}{0.5\textwidth}
      \begin{center}
        \includegraphics[width=0.5\textwidth]{c5.png}
      \end{center}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.
